---
title: "Differential Expression Analysis: Ewing Sarcoma vs Liposarcoma"
author: "Musfirah"
date: "17-07-2026"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE,
                      warning=FALSE,
                      message=FALSE,
                      fig.width=10,
                      fig.height=6)
```

```{r libraries}
library(DESeq2)
library(GEOquery)
library(ggplot2)
library(pheatmap)
library(enrichR)
library(Biobase)
```

```{r load-data, cache=TRUE}
options(timeout=300)
gse <- getGEO("GSE2553", GSEMatrix=TRUE,
              destdir="~/bioinformatics/data/")[[1]]
expr <- exprs(gse)
metadata <- pData(gse)
gene_info <- fData(gse)

# Select groups
ewing_idx <- which(metadata$source_name_ch1 == "Ewing Sarcoma")
lipo_idx <- which(metadata$source_name_ch1 == "Liposarcoma")
selected_idx <- c(ewing_idx, lipo_idx)
expr_subset <- expr[, selected_idx]
metadata_subset <- metadata[selected_idx,]
condition <- factor(c(rep("Ewing", length(ewing_idx)),
                      rep("Lipo", length(lipo_idx))))
```

```{r deseq2, cache=TRUE}
# Prepare count matrix
expr_counts <- round(2^expr_subset)
expr_counts[expr_counts < 0] <- 0
expr_counts <- matrix(as.integer(expr_counts),
                      nrow=nrow(expr_counts),
                      dimnames=dimnames(expr_counts))
expr_counts_clean <- expr_counts[complete.cases(expr_counts),]

# Create DESeq2 dataset
coldata <- data.frame(condition=condition,
                      row.names=colnames(expr_counts_clean))
dds <- DESeqDataSetFromMatrix(countData=expr_counts_clean,
                               colData=coldata,
                               design=~condition)
dds <- dds[rowSums(counts(dds)) > 10,]
dds <- DESeq(dds)
res <- results(dds, contrast=c("condition","Ewing","Lipo"))

# Process results
res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)
res_df$significance <- "Not Significant"
res_df$significance[res_df$padj < 0.05 &
                    res_df$log2FoldChange > 1] <- "Upregulated"
res_df$significance[res_df$padj < 0.05 &
                    res_df$log2FoldChange < -1] <- "Downregulated"

sig_degs <- subset(res_df, padj < 0.05 & abs(log2FoldChange) > 1)
top_up <- sig_degs[sig_degs$log2FoldChange > 0,]
top_up <- top_up[order(top_up$log2FoldChange, decreasing=TRUE),]
top_down <- sig_degs[sig_degs$log2FoldChange < 0,]
top_down <- top_down[order(top_down$log2FoldChange),]
```

## Abstract

This study performed differential expression analysis comparing
**Ewing Sarcoma** (n=`r length(ewing_idx)`) and **Liposarcoma**
(n=`r length(lipo_idx)`) samples from the GSE2553 dataset.
A total of **`r nrow(sig_degs)` significant DEGs** were identified
(padj < 0.05, |log2FC| > 1), including `r sum(sig_degs$log2FoldChange > 0)`
upregulated and `r sum(sig_degs$log2FoldChange < 0)` downregulated genes.
NKX2-2 was identified as a key Ewing Sarcoma biomarker.

## 1. Introduction

Soft tissue sarcomas represent a heterogeneous group of malignant tumors.
Ewing Sarcoma is characterized by the EWS-FLI1 fusion gene and primarily
affects children and young adults. Liposarcoma originates from adipose
tissue and is the most common soft tissue sarcoma in adults.
Understanding the molecular differences between these subtypes is crucial
for developing targeted therapies.

## 2. Methods

### 2.1 Data Source
GSE2553 dataset was downloaded from NCBI GEO using GEOquery.
Ewing Sarcoma (n=`r length(ewing_idx)`) and Liposarcoma
(n=`r length(lipo_idx)`) samples were selected for comparison.

### 2.2 Differential Expression Analysis
DESeq2 R package was used for differential expression analysis.
Genes with adjusted p-value < 0.05 and |log2 fold change| > 1
were considered significantly differentially expressed.

### 2.3 Pathway Enrichment
GO and KEGG pathway enrichment was performed using enrichR R package
on top 50 upregulated genes.

## 3. Results

### 3.1 Overview of Differential Expression

| Metric | Value |
|--------|-------|
| Total genes analyzed | `r nrow(res_df)` |
| Significant DEGs | `r nrow(sig_degs)` |
| Upregulated in Ewing | `r sum(sig_degs$log2FoldChange > 0)` |
| Downregulated in Ewing | `r sum(sig_degs$log2FoldChange < 0)` |

### 3.2 Volcano Plot

```{r volcano, fig.cap="Figure 1: Volcano plot of differential expression"}
ggplot(res_df, aes(x=log2FoldChange,
                    y=-log10(padj),
                    color=significance)) +
  geom_point(alpha=0.6, size=1.5) +
  scale_color_manual(values=c("Upregulated"="red",
                               "Downregulated"="blue",
                               "Not Significant"="grey")) +
  geom_vline(xintercept=c(-1,1), linetype="dashed") +
  geom_hline(yintercept=-log10(0.05), linetype="dashed") +
  labs(title="Volcano Plot: Ewing Sarcoma vs Liposarcoma",
       x="log2 Fold Change",
       y="-log10(adjusted p-value)",
       color="Expression") +
  theme_minimal()
```

### 3.3 Top Differentially Expressed Genes

```{r deg-table}
# Get gene symbols
top_up_sym <- gene_info[rownames(head(top_up,10)),"Symbol"]
top_down_sym <- gene_info[rownames(head(top_down,10)),"Symbol"]

deg_table <- data.frame(
  Gene = c(top_up_sym, top_down_sym),
  log2FoldChange = c(head(top_up,10)$log2FoldChange,
                     head(top_down,10)$log2FoldChange),
  padj = c(head(top_up,10)$padj,
            head(top_down,10)$padj),
  Direction = c(rep("Upregulated",10),
                rep("Downregulated",10))
)
knitr::kable(deg_table,
             caption="Table 1: Top 20 differentially expressed genes",
             digits=4)
```

### 3.4 Heatmap of Top DEGs

```{r heatmap, fig.cap="Figure 2: Heatmap of top DEGs", fig.height=10}
top_genes_idx <- c(rownames(head(top_up,10)),
                   rownames(head(top_down,10)))
top_expr <- expr_subset[top_genes_idx,]
gene_symbols <- gene_info[top_genes_idx,"Symbol"]
gene_symbols[gene_symbols==""] <- top_genes_idx[gene_symbols==""]
rownames(top_expr) <- gene_symbols

ann <- data.frame(Condition=condition,
                  row.names=colnames(top_expr))

pheatmap(top_expr,
         annotation_col=ann,
         show_colnames=FALSE,
         scale="row",
         main="Top DEGs: Ewing Sarcoma vs Liposarcoma",
         fontsize_row=8)
```

### 3.5 KEGG Pathway Enrichment

```{r kegg, fig.cap="Figure 3: KEGG pathway enrichment"}
up_symbols <- gene_info[rownames(top_up[1:50,]),"Symbol"]
up_symbols <- up_symbols[up_symbols != ""]

invisible(capture.output(
  enrich_results <- enrichr(up_symbols,
                            c("GO_Biological_Process_2023",
                              "KEGG_2021_Human"))
))

kegg_top <- enrich_results$KEGG_2021_Human[1:10,]
ggplot(kegg_top, aes(x=reorder(Term,-log10(P.value)),
                      y=-log10(P.value),
                      fill=-log10(P.value))) +
  geom_bar(stat="identity") +
  coord_flip() +
  scale_fill_gradient(low="steelblue", high="red") +
  labs(title="KEGG Pathways - Upregulated in Ewing Sarcoma",
       x="Pathway", y="-log10(P-value)") +
  theme_minimal() +
  theme(legend.position="none")
```

### 3.6 GO Biological Process Enrichment

```{r go, fig.cap="Figure 4: GO biological process enrichment"}
go_top <- enrich_results$GO_Biological_Process_2023[1:10,]
ggplot(go_top, aes(x=reorder(Term,-log10(P.value)),
                    y=-log10(P.value),
                    fill=-log10(P.value))) +
  geom_bar(stat="identity") +
  coord_flip() +
  scale_fill_gradient(low="steelblue", high="darkgreen") +
  labs(title="GO Biological Process - Upregulated in Ewing Sarcoma",
       x="GO Term", y="-log10(P-value)") +
  theme_minimal() +
  theme(legend.position="none")
```

## 4. Discussion

DESeq2 analysis identified `r nrow(sig_degs)` significant DEGs between
Ewing Sarcoma and Liposarcoma. **NKX2-2**, a transcription factor
directly regulated by the EWS-FLI1 fusion gene, was among the top
upregulated genes in Ewing Sarcoma, consistent with published literature.
**FABP4** was the most downregulated gene, reflecting the adipose origin
of Liposarcoma.

cAMP signaling pathway enrichment in Ewing Sarcoma upregulated genes
is consistent with the known role of EWS-FLI1 in activating cAMP-dependent
transcription. These findings provide molecular insights into the distinct
biology of these two sarcoma subtypes.

## 5. Conclusion

This DESeq2 analysis identified `r nrow(sig_degs)` significant DEGs
distinguishing Ewing Sarcoma from Liposarcoma. NKX2-2 and cAMP signaling
emerged as key molecular features of Ewing Sarcoma, providing potential
biomarkers and therapeutic targets for further investigation.

## References

1. GSE2553: NCBI GEO (PubMed ID: 16230383)
2. DESeq2: Love MI et al. Genome Biology (2014)
3. enrichR: Kuleshov MV et al. Nucleic Acids Research (2016)
4. R Core Team (2024)
