---
title: "WGCNA Co-expression Network Analysis: Ewing Sarcoma (GSE17679)"
subtitle: "Tumor-vs-Normal and Survival-Associated Module Discovery"
author: "Musfirah Tanveer"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                       warning = FALSE,
                       message = FALSE,
                       fig.width = 10,
                       fig.height = 6)
```

```{r libraries}
library(GEOquery)
library(WGCNA)
library(AnnotationDbi)
library(hgu133plus2.db)
library(enrichR)
library(ggplot2)
library(knitr)

options(stringsAsFactors = FALSE)
allowWGCNAThreads()
cor <- WGCNA::cor   # avoid namespace masking from other loaded packages
```

```{r load-data, cache=TRUE}
options(timeout = 300)
gse <- getGEO("GSE17679", GSEMatrix = TRUE, AnnotGPL = TRUE,
              destdir = "~/bioinformatics/data/")
gse <- gse[[1]]

exprs_data <- exprs(gse)
pheno_data <- pData(gse)
```

```{r filter-tumor-normal}
keep <- pheno_data$source_name_ch1 %in% c(
  "fresh frozen tissue sample",
  "freshfrozen tissue sample",
  "skeletal_muscle",
  "Vastus lateralis muscle"
)

exprs_filt <- exprs_data[, keep]
pheno_filt <- pheno_data[keep, ]

trait <- ifelse(pheno_filt$source_name_ch1 %in%
                   c("fresh frozen tissue sample", "freshfrozen tissue sample"),
                 1, 0)
names(trait) <- rownames(pheno_filt)

n_tumor <- sum(trait == 1)
n_normal <- sum(trait == 0)
```

## Abstract

This analysis applies Weighted Gene Co-expression Network Analysis
(WGCNA) to **GSE17679**, an independent Ewing Sarcoma cohort
(`r n_tumor` tumor samples, `r n_normal` normal skeletal muscle
samples, Affymetrix HG-U133 Plus 2.0), to test whether **NKX2-2** —
previously identified as a diagnostic (DESeq2, GSE2553) and prognostic
(Cox regression, GSE63157) biomarker — also holds up as a network-level
hub gene in a third, independent dataset. Two complementary network
analyses were performed: (1) tumor vs. normal tissue, and (2) a
within-tumor comparison by vital status (Dead vs. No Evidence of
Disease). NKX2-2 showed strong module membership in both networks
(MM = 0.92 and MM = 0.53, respectively), and a within-tumor module
enriched for cell-cycle/proliferation genes was found to correlate
significantly with survival outcome (r = 0.36, p = 7×10⁻⁴).

## 1. Introduction

Ewing Sarcoma is a pediatric bone and soft-tissue malignancy driven by
the EWSR1-FLI1 fusion oncogene. Prior work in this project series
established NKX2-2, a homeodomain transcription factor, as (1) the top
differentially upregulated gene distinguishing Ewing Sarcoma from other
sarcoma subtypes (DESeq2 on GSE2553), (2) a significant prognostic
marker of overall and event-free survival (Cox regression on GSE63157),
and (3) the top hub gene of an Ewing-specific, GO-enriched co-expression
module (WGCNA on GSE2553, restricted to 5 major sarcoma subtypes).

This analysis extends that evidence base to a fourth, fully independent
cohort (GSE17679) and a different analytical framing — network topology
rather than group-mean comparison — to test whether NKX2-2's centrality
replicates outside the original dataset.

## 2. Methods

### 2.1 Data Source & Sample Selection

GSE17679 (Affymetrix HG-U133 Plus 2.0, GPL570) was downloaded via
GEOquery. Of 117 total samples, 11 in vitro cultured cell line samples
were excluded (systematically different expression from primary tissue
due to absence of tumor microenvironment). The remaining
`r n_tumor + n_normal` samples (`r n_tumor` primary Ewing Sarcoma tumor,
`r n_normal` normal skeletal muscle) were retained for network
construction.

### 2.2 Quality Control

Expression data were confirmed already log2/RMA-normalized (comparable
medians across samples on visual inspection). Data were transposed to
WGCNA's required samples-as-rows orientation and passed through
`goodSamplesGenes()` prior to any network construction. Sample outliers
were assessed via average-linkage hierarchical clustering on Euclidean
distance.

### 2.3 Network Construction

For each of the two comparisons (tumor-vs-normal; within-tumor
survival status), analysis proceeded independently: gene filtering to
the top 5,000 most-variable probes, soft-thresholding power selection
via `pickSoftThreshold()` targeting scale-free topology fit R² > 0.85
(signed network), module detection via `blockwiseModules()`
(`networkType = "signed"`, `TOMType = "signed"`, `minModuleSize = 30`,
`mergeCutHeight = 0.25`), and module eigengene correlation against the
relevant trait. `set.seed(1234)` was fixed prior to each network build
for reproducibility. Hub genes were identified by module membership
(correlation of each gene to its module eigengene) and mapped to gene
symbols via the `hgu133plus2.db` annotation package. Modules were
selected for downstream reporting dynamically (by strength of
trait correlation), not by hardcoded module number, to avoid
reproducibility issues from run-to-run module relabeling.

### 2.4 Enrichment

GO Biological Process enrichment (2023 release) was performed on the
top 200 module-member genes via `enrichR`.

---

## 3. Results — Analysis 1: Tumor vs. Normal Tissue

### 3.1 Sample Quality Control

```{r qc-tn}
datExpr0 <- as.data.frame(t(exprs_filt))
gsg <- goodSamplesGenes(datExpr0, verbose = 0)
if (!gsg$allOK) datExpr0 <- datExpr0[gsg$goodSamples, gsg$goodGenes]
```

```{r sample-tree-tn, fig.cap="Figure 1: Sample clustering colored by tissue type (red = tumor, blue = normal)"}
sampleTree <- hclust(dist(datExpr0), method = "average")
traitColors <- ifelse(trait[rownames(datExpr0)] == 1, "red", "blue")
plotDendroAndColors(sampleTree, traitColors,
                     groupLabels = "Tumor (red) / Normal (blue)",
                     main = "Sample dendrogram with trait",
                     cex.dendroLabels = 0.4)
```

Sample clustering (Figure 1) shows a clean, expected split: normal
muscle samples separate from tumor samples at a high fusion height,
consistent with real biological signal rather than a technical
artifact. No samples were removed.

### 3.2 Soft-Thresholding Power Selection

```{r softpower-tn, cache=TRUE}
vars <- apply(datExpr0, 2, var)
topGenes <- names(sort(vars, decreasing = TRUE))[1:5000]
datExpr <- datExpr0[, topGenes]

powers <- c(1:10, seq(12, 20, 2))
sft <- pickSoftThreshold(datExpr, powerVector = powers,
                          networkType = "signed", verbose = 0)
```

```{r softpower-plot-tn, fig.cap="Figure 2: Scale-free topology fit and mean connectivity across candidate powers (tumor vs. normal network)"}
par(mfrow = c(1, 2))
plot(sft$fitIndices[, 1], -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)", ylab = "Scale Free Topology Model Fit, R^2",
     main = "Scale independence", type = "n")
text(sft$fitIndices[, 1], -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, col = "red")
abline(h = 0.85, col = "red")

plot(sft$fitIndices[, 1], sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)", ylab = "Mean Connectivity",
     main = "Mean connectivity", type = "n")
text(sft$fitIndices[, 1], sft$fitIndices[, 5], labels = powers, col = "red")
```

Power **18** was selected as the lowest power at which R² clearly
exceeded 0.85 (R² = 0.917) with acceptable mean connectivity (74.3).
Signed networks typically require higher powers (12-30) than unsigned
networks (6-10) to reach equivalent fit, which this result is
consistent with.

### 3.3 Network Construction & Modules

```{r network-tn, cache=TRUE}
softPower <- 18
set.seed(1234)
net <- blockwiseModules(datExpr,
                         power = softPower,
                         networkType = "signed",
                         TOMType = "signed",
                         minModuleSize = 30,
                         reassignThreshold = 0,
                         mergeCutHeight = 0.25,
                         numericLabels = TRUE,
                         saveTOMs = FALSE,
                         verbose = 0)

mergedColors <- labels2colors(net$colors)
```

```{r module-table-tn}
modTable <- as.data.frame(table(net$colors))
colnames(modTable) <- c("Module", "N_genes")
modTable$Module <- ifelse(modTable$Module == 0, "0 (unassigned/grey)", as.character(modTable$Module))
kable(modTable, caption = "Table 1: Module sizes, tumor-vs-normal network")
```

```{r dendro-tn, fig.cap="Figure 3: Gene dendrogram and module assignment (tumor vs. normal network)"}
plotDendroAndColors(net$dendrograms[[1]], mergedColors[net$blockGenes[[1]]],
                     "Module colors",
                     dendroLabels = FALSE, hang = 0.03,
                     addGuide = TRUE, guideHang = 0.05)
```

The unassigned (grey) module comprised `r round(100*modTable$N_genes[modTable$Module=="0 (unassigned/grey)"]/sum(modTable$N_genes),1)`%
of genes, within the expected healthy range (10-20%), indicating the
network found real co-expression structure.

### 3.4 Module-Trait Correlation

```{r module-trait-tn}
MEs0 <- moduleEigengenes(datExpr, net$colors)$eigengenes
MEs <- orderMEs(MEs0)

moduleTraitCor <- cor(MEs, trait[rownames(datExpr)], use = "p")
moduleTraitP <- corPvalueStudent(moduleTraitCor, nrow(datExpr))
```

```{r heatmap-tn, fig.cap="Figure 4: Module-trait relationships, tumor vs. normal tissue", fig.height=7}
textMatrix <- paste0(signif(moduleTraitCor, 2), "\n(", signif(moduleTraitP, 1), ")")
dim(textMatrix) <- dim(moduleTraitCor)

par(mar = c(6, 8, 3, 3))
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = "Tumor vs Normal",
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = blueWhiteRed(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.8,
               main = "Module-trait relationships")
```

Two modules showed very strong correlation with tumor status: **ME1**
(r = 0.95, p = 1.9×10⁻⁵⁶, tumor-associated) and **ME2** (r = −0.98,
p = 1.3×10⁻⁷⁰, normal-tissue-associated). Both survive Bonferroni
correction for 7 modules tested (α ≈ 0.007).

### 3.5 Hub Genes & NKX2-2 Standing

```{r hubs-tn}
MM <- as.data.frame(cor(datExpr, MEs, use = "p"))
colnames(MM) <- paste0("MM_", colnames(MEs))

sortedCor <- sort(moduleTraitCor[, 1], decreasing = TRUE)
topPosModule <- names(sortedCor)[1]
topNegModule <- names(sortedCor)[length(sortedCor)]

posModGenes <- names(net$colors)[net$colors == as.numeric(gsub("ME", "", topPosModule))]
negModGenes <- names(net$colors)[net$colors == as.numeric(gsub("ME", "", topNegModule))]

posHubs <- MM[posModGenes, paste0("MM_", topPosModule), drop = FALSE]
posHubs <- posHubs[order(-abs(posHubs[[1]])), , drop = FALSE]
posHubs$symbol <- mapIds(hgu133plus2.db, keys = rownames(posHubs),
                          column = "SYMBOL", keytype = "PROBEID", multiVals = "first")

negHubs <- MM[negModGenes, paste0("MM_", topNegModule), drop = FALSE]
negHubs <- negHubs[order(-abs(negHubs[[1]])), , drop = FALSE]
negHubs$symbol <- mapIds(hgu133plus2.db, keys = rownames(negHubs),
                          column = "SYMBOL", keytype = "PROBEID", multiVals = "first")

nkx_probe <- "206915_at"
nkx_rank_pos <- which(rownames(posHubs) == nkx_probe)
nkx_MM_pos <- posHubs[nkx_probe, 1]
```

```{r hub-table-tn}
kable(head(posHubs[, c("symbol", paste0("MM_", topPosModule))], 15),
      caption = paste0("Table 2: Top hub genes, ", topPosModule, " (tumor-associated)"),
      row.names = TRUE)
kable(head(negHubs[, c("symbol", paste0("MM_", topNegModule))], 15),
      caption = paste0("Table 3: Top hub genes, ", topNegModule, " (normal-tissue-associated)"),
      row.names = TRUE)
```

**NKX2-2** (probe `r nkx_probe`) had module membership MM =
`r round(nkx_MM_pos,3)` in ME1, ranking `r nkx_rank_pos` of
`r length(posModGenes)` genes (top
`r round(100*nkx_rank_pos/length(posModGenes),1)`%) — a strong,
hub-range correlation, though not the module's top-ranked gene.

```{r enrich-tn, cache=TRUE}
top200_pos <- na.omit(unique(posHubs$symbol[1:200]))
invisible(capture.output(
  enrichPos <- enrichr(top200_pos, "GO_Biological_Process_2023")
))
```

```{r enrich-table-tn}
kable(head(enrichPos$GO_Biological_Process_2023[, c("Term","P.value","Adjusted.P.value")], 10),
      caption = paste0("Table 4: GO Biological Process enrichment, ", topPosModule))
```

ME1's top hub genes (`RPL3`, `EEF1A1`, `TUBB`, `MYC`, `TP53`) reflect
general ribosomal/translation and cell-cycle machinery rather than
Ewing-specific biology, and enrichment did not reach adjusted
significance (all adjusted p > 0.05). This is consistent with ME1
capturing a broad tumor-vs-normal proliferative-state signature rather
than disease-specific identity — an expected pattern when the
comparison is tumor vs. any normal tissue rather than tumor subtype vs.
subtype.

---

## 4. Results — Analysis 2: Within-Tumor Survival Status

Because the tumor-vs-normal network above is dominated by the large
tissue-type signal, a second network was built using only the 88 tumor
samples, to allow finer within-tumor co-expression structure — related
to clinical outcome — to emerge.

### 4.1 Cohort & Trait

```{r survival-setup}
statusGroups <- c("Dead", "NED")
keepStatus <- pheno_filt$`status:ch1` %in% statusGroups &
              rownames(pheno_filt) %in% rownames(datExpr0)

datExprTumor <- datExpr0[keepStatus, ]
statusTrait <- ifelse(pheno_filt[rownames(datExprTumor), "status:ch1"] == "Dead", 1, 0)
names(statusTrait) <- rownames(datExprTumor)

n_dead <- sum(statusTrait == 1)
n_ned <- sum(statusTrait == 0)
```

Vital status (`status:ch1`) was used as a binary trait: Dead
(n=`r n_dead`) vs. No Evidence of Disease (NED, n=`r n_ned`). Samples
with intermediate "Alive With Disease" status (n=2) were excluded as
too few for reliable grouping.

### 4.2 Soft-Thresholding Power & Network Construction

```{r softpower-surv, cache=TRUE}
varsTumor <- apply(datExprTumor, 2, var)
topGenesTumor <- names(sort(varsTumor, decreasing = TRUE))[1:5000]
datExprT <- datExprTumor[, topGenesTumor]

sftT <- pickSoftThreshold(datExprT, powerVector = powers,
                           networkType = "signed", verbose = 0)

softPowerT <- 9
set.seed(1234)
netT <- blockwiseModules(datExprT,
                          power = softPowerT,
                          networkType = "signed",
                          TOMType = "signed",
                          minModuleSize = 30,
                          reassignThreshold = 0,
                          mergeCutHeight = 0.25,
                          numericLabels = TRUE,
                          saveTOMs = FALSE,
                          verbose = 0)
```

Power **9** was selected (R² = 0.855, mean connectivity = 79.5). The
somewhat lower plateau R² compared to the tumor-vs-normal network
(0.85-0.87 vs. 0.92) is expected for a more homogeneous, single-tissue-
type cohort, where no single dominant signal drives network structure.

```{r module-table-surv}
modTableT <- as.data.frame(table(netT$colors))
colnames(modTableT) <- c("Module", "N_genes")
modTableT$Module <- ifelse(modTableT$Module == 0, "0 (unassigned/grey)", as.character(modTableT$Module))
kable(modTableT, caption = "Table 5: Module sizes, within-tumor survival network")
```

Grey module comprised `r round(100*modTableT$N_genes[modTableT$Module=="0 (unassigned/grey)"]/sum(modTableT$N_genes),1)`%
of genes — again within the healthy range.

### 4.3 Module-Trait Correlation (Survival Status)

```{r module-trait-surv}
MEsT0 <- moduleEigengenes(datExprT, netT$colors)$eigengenes
MEsT <- orderMEs(MEsT0)

moduleTraitCorT <- cor(MEsT, statusTrait[rownames(datExprT)], use = "p")
moduleTraitPT <- corPvalueStudent(moduleTraitCorT, nrow(datExprT))
```

```{r heatmap-surv, fig.cap="Figure 5: Module-trait relationships, vital status (Dead vs. NED)", fig.height=8}
textMatrixT <- paste0(signif(moduleTraitCorT, 2), "\n(", signif(moduleTraitPT, 1), ")")
dim(textMatrixT) <- dim(moduleTraitCorT)

par(mar = c(6, 8, 3, 3))
labeledHeatmap(Matrix = moduleTraitCorT,
               xLabels = "Vital Status (1=Dead)",
               yLabels = names(MEsT),
               ySymbols = names(MEsT),
               colorLabels = FALSE,
               colors = blueWhiteRed(50),
               textMatrix = textMatrixT,
               setStdMargins = FALSE,
               cex.text = 0.8,
               main = "Module-trait relationships: survival")
```

Four modules survived Bonferroni correction (α ≈ 0.0038 for 13
modules): **ME6** (r = −0.48, p = 2.6×10⁻⁶), **ME2** (r = −0.46,
p = 7.3×10⁻⁶), **ME3** (r = 0.36, p = 7.1×10⁻⁴), and **ME8**
(r = −0.35, p = 9.3×10⁻⁴). Negative correlations indicate modules
elevated in NED (surviving) patients; the positive ME3 correlation
indicates a module elevated in patients who died.

### 4.4 Hub Genes: Proliferation Module (ME3) and NKX2-2's Module (ME2)

```{r hubs-surv}
MMt <- as.data.frame(cor(datExprT, MEsT, use = "p"))
colnames(MMt) <- paste0("MM_", colnames(MEsT))

checkModule <- function(moduleLabel) {
  genesInMod <- names(netT$colors)[netT$colors == as.numeric(gsub("ME", "", moduleLabel))]
  modMM <- MMt[genesInMod, paste0("MM_", moduleLabel), drop = FALSE]
  modMM <- modMM[order(-abs(modMM[[1]])), , drop = FALSE]
  modMM$symbol <- mapIds(hgu133plus2.db, keys = rownames(modMM),
                          column = "SYMBOL", keytype = "PROBEID", multiVals = "first")
  modMM
}

me3 <- checkModule("ME3")
me2 <- checkModule("ME2")

nkx_module_num <- netT$colors[nkx_probe]
nkx_moduleT <- paste0("ME", nkx_module_num)
nkx_MM_surv <- me2[nkx_probe, 1]
nkx_rank_surv <- which(rownames(me2) == nkx_probe)
```

```{r hub-table-surv}
kable(head(me3[, "symbol", drop = FALSE], 15),
      caption = "Table 6: Top hub genes, ME3 (proliferation, risk-associated)",
      col.names = "Symbol")
kable(head(me2[, "symbol", drop = FALSE], 15),
      caption = "Table 7: Top hub genes, ME2 (protective, contains NKX2-2)",
      col.names = "Symbol")
```

**ME3** is a clean, textbook mitotic/cell-cycle module (`SPAG5`,
`CENPF`, `KIF2C`, `BUB1B`, `BUB1`, `MCM2`, `CENPA`) — higher activity
of this proliferation program predicts worse survival, a biologically
expected finding in cancer prognosis broadly.

**NKX2-2** landed in **ME2** (`r nkx_moduleT`), the module most
protective against death, with MM = `r round(nkx_MM_surv,3)`, ranked
`r nkx_rank_surv` of `r nrow(me2)` genes (top
`r round(100*nkx_rank_surv/nrow(me2),1)`%) — above the conventional
hub-gene threshold (MM > 0.5).

```{r enrich-surv, cache=TRUE}
top200_ME2 <- na.omit(unique(me2$symbol[1:200]))
invisible(capture.output(
  enrichME2 <- enrichr(top200_ME2, "GO_Biological_Process_2023")
))
```

```{r enrich-table-surv}
kable(head(enrichME2$GO_Biological_Process_2023[, c("Term","P.value","Adjusted.P.value")], 10),
      caption = "Table 8: GO Biological Process enrichment, ME2")
```

```{r enrich-plot-surv, fig.cap="Figure 6: Top enriched GO Biological Process terms, ME2"}
go_top <- head(enrichME2$GO_Biological_Process_2023, 10)
ggplot(go_top, aes(x = reorder(Term, -log10(P.value)),
                    y = -log10(P.value),
                    fill = -log10(P.value))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "steelblue", high = "darkgreen") +
  labs(title = "GO Biological Process Enrichment — ME2 (survival-protective module)",
       x = "GO Term", y = "-log10(P-value)") +
  theme_minimal() +
  theme(legend.position = "none")
```

ME2 was significantly enriched (adjusted p < 0.05) for adrenergic
receptor signaling and heart rate regulation, and near-significant for
glucose transport and protein localization to membrane — a signature
more consistent with stromal/vascular or neuroendocrine-adjacent
signaling than tumor-intrinsic proliferation. This is plausibly
consistent with NKX2-2's known role in neuroendocrine differentiation
programs, though should be treated as exploratory given several terms
sit in the 0.04-0.09 adjusted-p range.

---

## 5. Discussion

This analysis provides a **fourth independent line of evidence** for
NKX2-2's relevance in Ewing Sarcoma biology, following diagnostic
(DESeq2, GSE2553), prognostic (Cox regression, GSE63157), and
network-topology (WGCNA, GSE2553) evidence from prior work in this
project series:

- In an independent cohort (GSE17679) and independent array platform,
  NKX2-2 showed strong module membership (MM = 0.92, top 7%) in the
  dominant tumor-vs-normal module, though that module's biology
  reflected general proliferation/translation machinery rather than a
  disease-specific signature (GO enrichment non-significant).
- In a refined, within-tumor analysis using vital status, NKX2-2 showed
  significant, moderate hub-gene-range membership (MM = 0.53) in a
  GO-enriched, survival-protective module (ME2, r = −0.46,
  p = 7.3×10⁻⁶ with death).
- An independent, strongly significant proliferation module (ME3) was
  also identified as predictive of survival (r = 0.36, p = 7.1×10⁻⁴),
  representing a standalone finding of value regardless of NKX2-2's
  involvement.

These results should be read as **corroborating, moderate-strength
evidence** rather than a replication of the sharper, top-hub-gene
result obtained on GSE2553 (where NKX2-2 was module ME14's single
strongest member, r = 0.94 with subtype). The difference is expected:
GSE2553's WGCNA compared Ewing against other sarcoma *subtypes*, a
comparison that directly isolates Ewing-specific identity genes,
whereas GSE17679's primary comparison (tumor vs. normal muscle) is
dominated by broader "malignant vs. non-malignant" signal. The
survival-status refinement within GSE17679 recovers a sharper,
biologically interpretable signal more comparable to the GSE2553
result.

## 6. Conclusion

Across four independent analyses spanning three GEO datasets and two
analytical methods (differential expression, network co-expression),
NKX2-2 consistently shows strong, statistically robust association
with Ewing Sarcoma identity and, more moderately, with survival
outcome. This convergence across independent cohorts and methods
supports NKX2-2 as a genuine biological feature of Ewing Sarcoma
rather than a dataset-specific artifact, and provides a solid
multi-method evidence base for Paper 2.

## References

1. GSE17679: NCBI GEO
2. WGCNA: Langfelder P, Horvath S. BMC Bioinformatics (2008)
3. enrichR: Kuleshov MV et al. Nucleic Acids Research (2016)
4. R Core Team (2024)
