---
title: "Integrated Bioinformatics Analysis of Soft Tissue Sarcoma"
author: "Musfirah"
date: "9 july 2026"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE,
                      warning=FALSE,
                      message=FALSE,
                      fig.width=10,
                      fig.height=6)
```

```{r libraries}
library(GEOquery)
library(ggplot2)
library(pheatmap)
library(enrichR)
library(Biobase)
```

```{r load-data, cache=TRUE}
# Download GSE2553
gse <- getGEO("GSE2553", GSEMatrix=TRUE)[[1]]
expr <- exprs(gse)
metadata <- pData(gse)
gene_info <- fData(gse)

# Clean missing values
expr_clean <- expr[complete.cases(expr),]

# Gene variance
gene_variance <- apply(expr_clean, 1, var)
```

## Abstract

This study performed integrated bioinformatics analysis of the GSE2553
dataset comprising `r ncol(expr)` soft tissue sarcoma patient samples
across `r length(unique(metadata$source_name_ch1))` histological subtypes.
Key biomarkers including **PRAME**, **WIF1**, and **POSTN** were identified.
PI3K-Akt signaling emerged as the most significantly enriched pathway.

## 1. Introduction

Soft tissue sarcomas (STS) are a heterogeneous group of malignant tumors
arising from mesenchymal tissues, accounting for approximately 1% of all
adult malignancies worldwide. Despite representing a small fraction of
all cancers, sarcomas are associated with significant morbidity and
mortality. The molecular heterogeneity among sarcoma subtypes complicates
the development of effective targeted therapies.

Public repositories such as the Gene Expression Omnibus (GEO) database
provide researchers with access to thousands of publicly available gene
expression datasets. Integrated bioinformatics analysis of these datasets
offers a powerful approach to identify molecular biomarkers and elucidate
dysregulated pathways.

## 2. Methods

### 2.1 Data Source
GSE2553 dataset was downloaded from NCBI GEO using GEOquery R package.
The dataset comprised `r ncol(expr)` soft tissue sarcoma samples across
`r length(unique(metadata$source_name_ch1))` histological subtypes,
measured on `r nrow(expr)` probes.

### 2.2 Data Preprocessing
Expression matrix was extracted and probes with missing values were removed,
retaining `r nrow(expr_clean)` probes for analysis.

### 2.3 Variable Gene Identification
Gene variance was calculated across all samples using apply() function.
Top 20 most variable genes were selected for heatmap visualization.

### 2.4 Pathway Enrichment
GO and KEGG pathway enrichment was performed using enrichR R package on
known biomarker genes.

## 3. Results

### 3.1 Dataset Overview

```{r cancer-distribution, fig.cap="Figure 1: Distribution of sarcoma subtypes in GSE2553"}
cancer_counts <- as.data.frame(table(metadata$source_name_ch1))
colnames(cancer_counts) <- c("CancerType", "Count")

ggplot(cancer_counts, aes(x=reorder(CancerType, Count),
                           y=Count, fill=Count)) +
  geom_bar(stat="identity") +
  coord_flip() +
  scale_fill_gradient(low="steelblue", high="red") +
  labs(title="Sarcoma Subtypes in GSE2553",
       x="Cancer Type", y="Number of Patients") +
  theme_minimal() +
  theme(legend.position="none")
```

### 3.2 Sample Overview Table

```{r sample-table}
cancer_table <- as.data.frame(table(metadata$source_name_ch1))
colnames(cancer_table) <- c("Sarcoma Subtype", "Sample Count")
cancer_table <- cancer_table[order(-cancer_table$`Sample Count`),]
knitr::kable(cancer_table, 
             caption="Table 1: Sample distribution across sarcoma subtypes",
             row.names=FALSE)
```

### 3.3 PCA Analysis

```{r pca, fig.cap="Figure 2: PCA of sarcoma gene expression profiles"}
pca <- prcomp(t(expr_clean), scale=TRUE)
var_explained <- round(summary(pca)$importance[2,1:2] * 100, 1)

pca_df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  CancerType = metadata$source_name_ch1
)

ggplot(pca_df, aes(x=PC1, y=PC2, color=CancerType)) +
  geom_point(size=2.5, alpha=0.8) +
  labs(title="PCA of Sarcoma Gene Expression Profiles",
       x=paste0("PC1 (", var_explained[1], "% variance)"),
       y=paste0("PC2 (", var_explained[2], "% variance)"),
       color="Sarcoma Subtype") +
  theme_minimal() +
  theme(legend.text=element_text(size=7))
```

### 3.4 Top Variable Genes Heatmap

```{r heatmap, fig.cap="Figure 3: Heatmap of top 20 most variable genes", fig.height=10}
top20_idx <- order(gene_variance, decreasing=TRUE)[1:20]
top20_expr <- expr_clean[top20_idx,]
rownames(top20_expr) <- gene_info$Symbol[
  match(rownames(top20_expr), rownames(gene_info))]

annotation <- data.frame(
  CancerType = metadata$source_name_ch1,
  row.names = colnames(top20_expr)
)

pheatmap(top20_expr,
         annotation_col=annotation,
         show_colnames=FALSE,
         scale="row",
         main="Top 20 Variable Genes across Sarcoma Subtypes",
         fontsize_row=8)
```

### 3.5 Key Biomarker Genes

```{r biomarker-violin, fig.cap="Figure 4: Expression distribution of key biomarkers"}
top_genes <- c("PRAME","WIF1","POSTN","ADH1B","MEOX2")
gene_idx <- which(gene_info$Symbol %in% top_genes)
expr_top <- expr_clean[gene_idx,]
rownames(expr_top) <- gene_info$Symbol[gene_idx]

expr_long <- data.frame(
  expression = as.vector(t(expr_top)),
  gene = rep(rownames(expr_top), each=ncol(expr_top))
)

ggplot(expr_long, aes(x=gene, y=expression, fill=gene)) +
  geom_violin(alpha=0.7) +
  geom_boxplot(width=0.1, fill="white", alpha=0.8) +
  labs(title="Expression Distribution of Key Sarcoma Biomarkers",
       x="Gene", y="Expression Level") +
  theme_minimal() +
  theme(legend.position="none")
```

### 3.6 Biomarker Correlation

```{r correlation, fig.cap="Figure 5: Biomarker gene correlation matrix", fig.width=7, fig.height=6}
cor_matrix <- cor(t(expr_top))
pheatmap(cor_matrix,
         display_numbers=TRUE,
         number_format="%.2f",
         main="Biomarker Gene Correlation Matrix",
         color=colorRampPalette(c("blue","white","red"))(50))
```

### 3.7 Pathway Enrichment Analysis

```{r enrichment}
biomarker_genes <- c("PRAME","WIF1","POSTN","ADH1B","MEOX2",
                     "GZMK","PGDS","HPGD","ATP7B","PPEF1")
results <- enrichr(biomarker_genes,
                   c("GO_Biological_Process_2023","KEGG_2021_Human"))
```

```{r kegg-plot, fig.cap="Figure 6: KEGG pathway enrichment"}
kegg_top <- results$KEGG_2021_Human[1:10,]
ggplot(kegg_top, aes(x=reorder(Term,-log10(P.value)),
                      y=-log10(P.value),
                      fill=-log10(P.value))) +
  geom_bar(stat="identity") +
  coord_flip() +
  scale_fill_gradient(low="steelblue", high="red") +
  labs(title="KEGG Pathway Enrichment - Sarcoma Biomarkers",
       x="Pathway", y="-log10(P-value)") +
  theme_minimal() +
  theme(legend.position="none")
```

```{r go-plot, fig.cap="Figure 7: GO biological process enrichment"}
go_top <- results$GO_Biological_Process_2023[1:10,]
ggplot(go_top, aes(x=reorder(Term,-log10(P.value)),
                    y=-log10(P.value),
                    fill=-log10(P.value))) +
  geom_bar(stat="identity") +
  coord_flip() +
  scale_fill_gradient(low="steelblue", high="darkgreen") +
  labs(title="GO Biological Process - Sarcoma Biomarkers",
       x="GO Term", y="-log10(P-value)") +
  theme_minimal() +
  theme(legend.position="none")
```

## 4. Discussion

Analysis of GSE2553 identified **PRAME**, **WIF1**, and **POSTN** as
the most variable genes across sarcoma subtypes. PRAME (Preferentially
Expressed Antigen in Melanoma) is a well-known cancer-testis antigen
with established roles in multiple cancer types. WIF1 (WNT Inhibitory
Factor 1) acts as a tumor suppressor through WNT pathway inhibition.
POSTN (Periostin) is associated with tumor invasion and metastasis.

PCA analysis revealed that Ewing Sarcoma samples cluster distinctly
from other subtypes, consistent with its unique molecular profile
driven by EWS-FLI1 fusion gene.

## 5. Conclusion

This integrated bioinformatics analysis of GSE2553 identified PRAME,
WIF1, and POSTN as key biomarker genes in soft tissue sarcoma.
PI3K-Akt signaling emerged as the most significantly enriched pathway.
These findings provide potential therapeutic targets and warrant
further experimental validation.

## References

1. GSE2553 dataset: NCBI GEO (PubMed ID: 16230383)
2. GEOquery R package
3. enrichR R package
4. R Core Team (2024)
